Updated Criteria for the Diagnostic Procedure for Parkinson’s Disease Dementia on Level I and their Validity in Deep Brain Stimulation Cohort

### 2025-05-13 _manuscript/manuscript.qmd
Final introduction refinement (JM approved).

Parkinson’s disease (PD) is a neurodegenerative disorder defined by a gradual onset of motor impairment, including rigidity, bradykinesia, postural instability and resting tremor. Additionally, patients frequently suffer from a range of non-motor impairments [@postuma2015], particularly cognitive decline. This state might result in Parkinson’s disease dementia (PDD) in a subset of patients [@meireles2012].

Despite its clinical relevance, the diagnosis of PDD remains complex. The original criteria formalized in 2007 by the International Parkinson and Movement Disorder Society (MDS) [@dubois2007] were heavily influenced by frameworks established primarily for Alzheimer’s disease (AD). Although these criteria provided a valuable base, they lacked the specificity to detect the specific cognitive aspects of PDD [@emre2007; @yamashita2023].

In order to increase diagnostic accuracy, the MDS also developed a levelled system for PDD detection. Level I criteria are designed primarily for clinical use, consisting of brief cognitive assessments and relying on clinical judgement. In contrast, Level II criteria use comprehensive neuropsychological testing across cognitive domains to provide greater diagnostic specificity [@emre2007].

One specific feature of the original Level I criteria was the presence of an algorithm allowing flexibility in test selection. The criteria included MMSE pentagons for evaluation of visuospatial ability, three-word recall for memory assessment,  months reversed or seven backwards for attention assessment and lexical fluency or clock drawing for executive function assessment [@dubois2007].

According to a recent meta-nalysis, the PDD prevalence estimates vary widely, ranging from 14% up to 55%, depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients` sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline [@oh2016; @rana2012]. Currently, efforts are focused on refining the PDD diagnostic framework to improve its consistency and applicability in both research and clinical contexts [@kulisevsky2024].

The present study evaluates the diagnostic concordance between the original Level I PDD criteria, as established by the MDS Task Force [@emre2007; @dubois2007] and criteria inspired by the recent call for change [@kulisevsky2024] within a cohort of PD patients considered for DBS. Furthermore, both sets of criteria are compared to PDD diagnosed on Level II. The study aims to address the following research objectives (RO): (RO1) To estimate the Parkinson’s disease dementia (PDD) rates and evaluate the diagnostic variability and concordance across different PDD criteria. (RO2) To identify specific diagnostic components contributing to PDD classification variability across the applied criteria.

@@

This study retrospectively analyzed clinical data from a cohort of patients with PD at the General University Hospital in Prague. All patients were diagnosed with idiopathic PD by a movement disorder specialist according to the MDS Clinical Diagnostic Criteria for PD [@postuma2015]. Clinical records spanning `r dates["min"]` to `r dates["max"]` were examined. All participants underwent neuropsychological evaluation conducted by a trained clinical psychologist (OB) as part of standard preoperative cognitive assessments for DBS eligibility at the General University Hospital in Prague.

@@

Our results showed a wide range in estimated PDD prevalence across the total of 68 algorithms, ranging from 2.00% to 16.75%. Estimates reached lower rates when using solely FAQ item 9 in comparison with the full FAQ scale. This discrepancy highlights the diagnostic importance of IADL assessment. Our overall PDD rates were consistently lower compared to previous studies regarding PDD among PD patients, demonstrating wide variability based on various criteria used. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigation found even higher rates, reaching up to 30% [@aarsland2011]. A recent complex meta-analysis synthesizing global data placed the rate at 26.30% [@sousa2022]. **The discrepancy is likely attributable to our DBS-specific cohort, which differs in important characteristics, such as the younger age of the patients [@rana2012]. / Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in sample characteristics, diagnostic criteria, and methodology. [^3]**

[^3]: **MM to JM: Which sentence do you like better?**

@@

Estimates relating to RO1.1 and RO1.3 cannot be safely generalized beyond a population of PD patients that are candidates for DBS due to the systematic differences between DBS candidates pool and general PD population (such as the lower age of DBS candidates compared to the general PD population). On the other hand, the estimates relating to RO1.4 (and to a lesser degree to RO1.2[^3]) may not be substantially influenced by the sample at hand as the primary source of their variance might come from variability in measures employed (e.g., MMSE vs MoCA to assess global cognitive performance) rather than variability in patients' performance. Assuming that there is no substantial Differential Item Functioning for DBS candidates compared to a broader population of patients with PD, the estimates relating to RO1.4 can be cautiously generalized beyond the current sample.

[^3]: Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events

>>>>

Parkinson’s disease (PD) is a neurodegenerative disorder. It is primarily defined by a gradual onset of motor impairment, including rigidity, bradykinesia, postural instability and resting tremor. Additionally, patients frequently suffer from a range of non-motor impairments [@postuma2015], particularly cognitive decline. This state might result in Parkinson’s disease dementia (PDD) in a subset of patients [@meireles2012].

According to a recent meta-analysis, approximately one-quarter of PD patients is likely to be diagnosed with PDD [@sousa2022]. However, reported PDD rate estimates vary widely, ranging from 14% up to 55%, depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients' sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline and PDD [@oh2016; @rana2012].

Despite the clinical relevance of PDD, its diagnosis remains complex. A milestone in research of PDD was the publication of diagnostic criteria established in 2007 by the International Parkinson and Movement Disorder Society (MDS) [@dubois2007]. In these citeria, the MDS introduced a two-levelled system for PDD detection. Level I consists of brief cognitive assessments, while Level II involves comprehensive neuropsychological testing across cognitive domains [@emre2007].

The original Level I algorithm included eight conditions that had to be satisfied simultaneously in order to diagnose probable PDD. These included: 1) diagnosis of PD proposed by the Queen Square Brain Bank; 2) PD onset prior to the PDD emergence; 3) evidence of global cognitive impairment (MMSE score \< 26 points); 4) cognitive deficit interference with the IADL (assessed by the pill questionnaire or caregiver interview); 5) impairment in at least two cognitive domains, namely memory, attention, visuo-constructive abilities and executive functions; 6) there was absence of Major Depressive Disorder; 7) absence of delirium; and 8) exclusion of other abnormalities and potential causes of dementia [@dubois2007].

Currently, efforts are focused on refining this PDD diagnostic framework. A recent call for a change pinpoints limitations regarding the original criteria and suggest various updates to enhance their utility [@kulisevsky2024]. Proposed suggestions include replacement of Mini Mental State Examination (MMSE) by Montreal Cognitive Assessment (MoCA), which is more sensitive to PD specific cognitive impairment; expansion of instrumental activities of daily living (IADL) evaluation; inclusion of language assessment; recognition of anxiety as one the neuropsychiatric symptoms relevant in PDD; and intergration of biomarkers.

In the light of these proposals, the current study aims to evaluate the diagnostic concordance between the original MDS Level I PDD criteria [@emre2007; @dubois2007] and a modified framework based on the recent call for change [@kulisevsky2024]. Furthermore, both Level I diagnostic approaches are compared to PDD diagnosed on Level II. The study aims to address the following research objectives (RO): (RO1) To estimate the PDD rate and evaluate the diagnostic variability and concordance across different PDD criteria. (RO2) To identify specific diagnostic components contributing to PDD classification variability across the applied criteria.

@@

This study retrospectively analyzed clinical data from a cohort of patients with PD at the General University Hospital in Prague. All patients were diagnosed with idiopathic PD by a movement disorder specialist according to the MDS Clinical Diagnostic Criteria for PD [@postuma2015]. Clinical records spanning `r dates["min"]` to `r dates["max"]` were examined. All participants were candidates for for Deep Brain Stimulation (DBS) treatment and underwent neuropsychological evaluation conducted by a trained clinical psychologist (OB) as part of standard preoperative assessments forDBS eligibility at the General University Hospital in Prague.

@@

Our results showed a wide range in estimated PDD prevalence across the total of 68 algorithms, ranging from 2.00% to 16.75%. Estimates reached lower rates when using solely FAQ item 9 in comparison with the full FAQ scale. This discrepancy highlights the diagnostic importance of IADL assessment. Our overall PDD rates were consistently lower compared to previous studies regarding PDD among PD patients, demonstrating wide variability based on various criteria used. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigation found even higher rates, reaching up to 30% [@aarsland2011]. A recent complex meta-analysis synthesizing global data placed the rate at 26.30% [@sousa2022]. **The discrepancy is likely attributable to our DBS-specific cohort, which differs in important characteristics, such as the younger age of the patients [@rana2012]. / Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in sample characteristics, diagnostic criteria, and methodology.** [^3]

[^3]: **Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events**

@@

Estimates relating to RO1.1 and RO1.3 cannot be safely generalized beyond a population of PD patients that are candidates for DBS due to the systematic differences between DBS candidates pool and general PD population (such as the lower age of DBS candidates compared to the general PD population). On the other hand, the estimates relating to RO1.4 (and to a lesser degree to RO1.2[^4]) may not be substantially influenced by the sample at hand as the primary source of their variance might come from variability in measures employed (e.g., MMSE vs MoCA to assess global cognitive performance) rather than variability in patients' performance. Assuming that there is no substantial Differential Item Functioning for DBS candidates compared to a broader population of patients with PD, the estimates relating to RO1.4 can be cautiously generalized beyond the current sample.

[^4]: Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events
